---
title: "The number, taxonomic origins and relative abundance of MAGs"
author: "Julius Eyiuche Nweze"
date: "2026-04-23"
output:
  rmarkdown::html_document:
    code_folding: show
    dev: png
    df_print: paged
    fig_caption: yes
    highlight: pygments
    keep_md: yes
    number_sections: no
    theme: flatly
    toc: yes
    toc_depth: 5
    toc_float: yes
editor_options: 
  chunk_output_type: console
---
<!--
Title          : 02_MAG_abundance.Rmd
Description    : Join CoverM MAG relative abundance with sample metadata, anvi'o bin summaries and GTDB-Tk novelty; plot unmapped reads, MAG counts, phylum- and family-level composition, and genus- and species-level novelty
Author         : Julius Eyiuche Nweze
Date           : 2026/04/23
Usage          : rmarkdown::render("02_MAG_abundance.Rmd")
-->


```{r libraries, include=F}
repo <- "http://cran.wu.ac.at"
lib.loc <- Sys.getenv("R_LIBS_USER")

update.packages(
    lib.loc, 
    repos = repo,
    ask = FALSE
)

.cran_libs <- c(
"devtools", # The tools required to efficiently manage other packages
"data.table", # Optimized for fast reading of large files (fread)
"readr", # For reading .csv files efficiently
"tibble", # For tibble manipulations (data frames with extra features)
"readODS", # Read ODS Files
"tidyverse", # Helps to transform and better present data. 
"kableExtra",
"FSA",
"car",
"rstatix", 
"vegan", 
"ggplot2", # For creating declarative and customizable graphics
"magrittr", # Forward and backward “pipe”-like operator
"dplyr" # A grammar of data manipulation
)

.inst <- .cran_libs %in% installed.packages()
if (any(!.inst)) {
   install.packages(.cran_libs[!.inst],
                    repos = repo,
                    lib = lib.loc)
}

.bioc_libs <- c(
  #"multtest", #Resampling-based multiple hypothesis testing
)

.bioc_inst <- .bioc_libs %in% installed.packages()
if (any(!.bioc_inst)) {
   if (!requireNamespace("BiocManager", quietly = TRUE))
   install.packages("BiocManager")
   BiocManager::install(ask = F, lib = lib.loc)  # upgrade bioC packages
   BiocManager::install(.bioc_libs[!.bioc_inst], ask = F, lib = lib.loc)
}

.local_libs <- c()

.inst <- names(.local_libs) %in% installed.packages()
if (any(!.inst)) {
   install.packages(paste0("~/R/", .local_libs[!.inst]) ,repos = NULL, type = "source", lib = lib.loc)
}

.github_libs <- c(
  "wilkelab/ggtext", # Improved text rendering support for 'ggplot2' 
  "ACCLAB/dabestr" # Data Analysis using Bootstrap-Coupled Estimation 
)

.github_lib_names <- stringr::str_replace(.github_libs, ".*/(.*)$", "\\1")

.github_inst <- .github_lib_names %in% installed.packages()
if (any(!.github_inst)) {
  devtools::install_github(.github_libs[!.github_inst],
                           lib = lib.loc,
                           dependencies = TRUE)
}

# Load packages into session, and print package version
(loaded.libs <- sapply(c(.cran_libs, .bioc_libs, names(.local_libs), .github_lib_names), require, character.only = TRUE))
if (!all(loaded.libs)) {stop(paste("Package(s):", names(loaded.libs[loaded.libs == FALSE]), "could not be loaded"))}
sapply(c(.cran_libs, .bioc_libs, names(.local_libs), .github_lib_names), packageVersion)
```


**Load Unmapped metagenome abundance and metadata**
```{r load data, cache = T}
# --- Load Unmapped MAG Abundance Data ---
read_ods("Exp3_MG_coverm_merged.ods", sheet = "Exp3_MG_coverm_merged",  col_names = TRUE) %>%
  filter(MAGs == "unmapped") ->
  Exp3_MG_unmap

# --- Load Metadata ---
read_ods("Metadata_MG.ods", sheet = "Metadata_MG",  col_names = TRUE) ->
  Metadata_MG_unmap

# --- Merge Abundance Data with Metadata ---
Abs_metadata_MG_unmap <- Metadata_MG_unmap %>% 
  dplyr::right_join(Exp3_MG_unmap, by=c("SampleID"), multiple = "all", relationship = "many-to-many") 

# --- Save Merged Data to CSV ---
write_csv(Abs_metadata_MG_unmap, "Unmapped_MAGs_abundance_metadata_MG.csv")


```

**Plot unmapped reads**
```{r load unmapped data, cache = T}
# --- Define Plant Colors ---
plant = c(Typha ="#8B4513", Juncus = "#53868B")

# --- Calculate Mean Relative Abundance of Unmapped MAGs ---
MAG_abd_unmap <- Abs_metadata_MG_unmap %>%
  subset(Relative_abundance > 0) %>%                              # Keep only positive abundances
  mutate(
    `Water type` = factor(`Water type`, levels = c("OSPW", "ROW")), # Order water type factor
    `Plant species` = factor(`Plant species`, levels = c("Typha", "Juncus")), # Order plant species factor
    Time = factor(Time, levels = c("D-0", "D6", "D19", "D40", "D60", "D89")) # Order time points
  ) %>%
  # Average across replicates
  group_by(MAGs, SampleID, `Plant species`, `Water type`, Time, Replicates) %>%
  mutate(DE = mean(Relative_abundance), .groups = "drop") %>%
  # Summarize across replicates for plotting
  group_by(`Plant species`, `Water type`, Time) %>%
  summarise(DE_mean = mean(DE), .groups = "drop") 

# --- Plot Unmapped MAGs Relative Abundance ---
MAG_abd_unmap %>%
  ggplot(aes(x = `Water type`, y = DE_mean, fill = `Plant species`)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  xlab("Plant species") +
  ylab("Unmapped reads to MAGs (%)") +
  theme_bw() +
  geom_text(aes(label = round(DE_mean, 1)), position = position_dodge(width = 0.9), angle = 90, hjust = 0.05, size = 8) +
  scale_fill_manual(values = plant) +
  facet_wrap(~ Time, ncol = 3) +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    strip.text.x = element_text(size = 26),
    strip.text.y = element_text(size = 26),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(size = 24)
  )

# --- Save Plot as SVG ---
ggsave("Figures/Percentage_of_Unmapped_reads.svg", width = 45, height = 55, units = "cm")

```


**Load MAGs abundance and metadata**
```{r load MAGs data, cache = T}

# --- Load MAG Abundance Data ---
read_ods("Exp3_MG_coverm_merged.ods", sheet = "Exp3_MG_coverm_merged",  col_names = TRUE) ->
  Exp3_MG

# --- Load Metadata ---
read_ods("Metadata_MG.ods", sheet = "Metadata_MG",  col_names = TRUE) ->
  Metadata_MG

# --- Load Bin Summary Metadata ---
read_ods("bins_summary.ods", sheet = "bins_summary",  col_names = TRUE) ->
  anvi_sum_MG

# --- Load Curated Novelty/Taxonomy Metadata ---
read_ods("bins_summary.ods", sheet = "Novelty", col_names = TRUE) %>%
  dplyr::select(MAGs, Domain, Phylum, Class, Order, Family, Genus, Species) ->
  novelty_taxonomy_MG

# --- Replace Older Bin-Summary Taxonomy With Curated Novelty Taxonomy ---
anvi_sum_MG <- anvi_sum_MG %>%
  dplyr::select(-Domain, -Phylum, -Class, -Order, -Family, -Genus, -Species) %>%
  dplyr::left_join(novelty_taxonomy_MG, by = "MAGs")

# --- Merge Abundance Data with Sample Metadata ---
Abs_metadata_MG <- Metadata_MG %>% 
  dplyr::right_join(Exp3_MG, by=c("SampleID"), multiple = "all", relationship = "many-to-many") 

# --- Merge with Bin Summary Metadata ---
Abs_metadata_MG <- Abs_metadata_MG %>% 
  dplyr::right_join(anvi_sum_MG, by=c("MAGs"), multiple = "all", relationship = "many-to-many") 

# --- Save Merged Data to CSV ---
write_csv(Abs_metadata_MG, "MAGs_abundance_metadata_MG.csv")


```


**Calculate relative abundance of MAGs in metagenome for ITOL**
```{r load calcuate ITOL, cache = T}
# --- Prepare MAG Abundance Table for ITOL ---

Abs_metadata_MG %>%
  # Optional: Filter for a specific plant species
  # subset(`Plant species` == "Typha") %>%
  
  # --- Factorize Columns for Consistent Ordering ---
  mutate(`Water type` = factor(`Water type`, levels = c("OSPW", "ROW"))) %>%
  mutate(`Plant species` = factor(`Plant species`, levels = c("Typha", "Juncus"))) %>%
  mutate(Time = factor(Time, levels = c("D-0", "D6", "D19", "D40", "D60", "D89"))) %>%
  
  # --- Average Across Replicates ---
  group_by(MAGs, SampleID, `Plant species`, `Water type`, Time, Replicates) %>%
  mutate(DE = mean(Relative_abundance)) %>%
  
  # --- Log-transform Abundance (optional, useful for visualization) ---
  mutate(DE_log = log10(DE + 1)) %>%
  
  # --- Reshape Data to Wide Format for ITOL ---
  reshape2::dcast(MAGs ~ `Plant species` + `Water type` + Time,
                  value.var = "DE",
                  fun.aggregate = mean) %>%
  
  # --- Save as CSV for ITOL ---
  write.csv("ITOL/Exp_MG_abundance.csv")

```

**Number of MAGs**
```{r load n MAGs, cache = T}
# --- Define Plant Colors for Plot ---
plant = c(Typha ="#8B4513", Juncus = "#53868B")

# --- Calculate Number of MAGs per Plant, Water Type, and Time ---
MAG_abd_no <- Abs_metadata_MG %>%
  subset(Relative_abundance > 0) %>%                       # Only consider MAGs with positive abundance
  mutate(
    `Water type` = factor(`Water type`, levels = c("OSPW", "ROW")),   # Order water types
    `Plant species` = factor(`Plant species`, levels = c("Typha", "Juncus")), # Order plant species
    Time = factor(Time, levels = c("D-0", "D6", "D19", "D40", "D60", "D89")) # Order time points
  ) %>%
  # Average across replicates
  group_by(MAGs, SampleID, `Plant species`, `Water type`, Time, Replicates) %>%
  mutate(DE = mean(Relative_abundance), .groups = "drop") %>%
  # Count unique MAGs per plant, water type, and time
  group_by(`Plant species`, `Water type`, Time) %>%
  summarise(count = n_distinct(MAGs), .groups = "drop") 

# --- Plot Number of MAGs ---
MAG_abd_no %>%
  ggplot(aes(x = Time, y = count, fill = `Plant species`)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_text(aes(label = count), position = position_dodge(width = 0.9), angle = 90, hjust = 0.05, size = 8) +
  xlab("Plant species") +
  ylab("Number of MAGs") +
  facet_wrap(~`Water type`) +
  theme_bw() +
  scale_fill_manual(values = plant) +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    strip.text.x = element_text(size = 26),
    strip.text.y = element_text(size = 26),
    legend.text = element_text(colour = "black", size = 24),
    axis.text.x = element_text(
      size = 24, hjust = 1, vjust = 0.5,
      angle = 90, colour = "black"
    ),
    legend.title = element_text(size = 24)
  )

# --- Save Plot ---
ggsave("Figures/Number_of_MAGs_Plant_Time.svg", width = 45, height = 45, units = "cm")

```


**Number of MAGs at phylum level**
```{r load n MAGs phylum, cache = T}
# --- Define Colors for Plants ---
plant = c(Typha ="#8B4513", Juncus = "#53868B")

# --- Prepare Data: Number of MAGs at Phylum Level ---
MAG_abd_no <- Abs_metadata_MG %>%
  # Keep only MAGs with non-zero abundance
  subset(Relative_abundance > 0) %>%
  
  # Factorize columns for ordering in plots
  mutate(
    `Water type` = factor(`Water type`, levels = c("OSPW", "ROW")),
    `Plant species` = factor(`Plant species`, levels = c("Typha", "Juncus")),
    Time = factor(Time, levels = c("D-0", "D6", "D19", "D40", "D60", "D89"))
  ) %>%
  
  # --- Average Across Replicates ---
  group_by(MAGs, SampleID, `Plant species`, `Water type`, Time, Replicates) %>%
  mutate(DE = mean(Relative_abundance), .groups = "drop") %>%
  
  # --- Count MAGs per Phylum, Plant, Water Type, and Time ---
  group_by(`Plant species`, `Water type`, Time, Phylum) %>%
  summarise(count = n_distinct(MAGs), .groups = "drop") 

# --- Plot Number of MAGs ---
MAG_abd_no %>%
  ggplot(aes(x = Phylum, y = count, fill = `Plant species`)) +
  
  # Bar plot
  geom_bar(stat = "identity", position = position_dodge()) +
  
  # Add counts as text labels
  geom_text(aes(label = count), position = position_dodge(width = 0.9),
            angle = 90, hjust = 0.05, size = 4) +
  
  # Axis labels
  xlab("Plant species") +
  ylab("Number of MAGs") +
  
  # Theme and color settings
  theme_bw() +
  scale_fill_manual(values = plant) +
  
  # Facet by Water Type and Time
  facet_wrap(`Water type` ~ Time, ncol = 3) +
  
  # Customize text and legend
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    strip.text.x = element_text(size = 26),
    strip.text.y = element_text(size = 26),
    legend.text = element_text(colour = "black", size = 24),
    axis.text.x = element_text(
      size = 24, hjust = 1, vjust = 0.5,
      angle = 90, colour = "black"
    ),
    legend.title = element_text(size = 24)
  )

# --- Save Plot ---
ggsave("Figures/Number_of_MAGs_Phylum.svg", width = 55, height = 65, units = "cm")

```


** Number of MAGs at family level**
```{r load n MAGs family, cache = T}

# --- Define Colors for Plants ---
plant = c(Typha ="#8B4513", Juncus = "#53868B")

# --- Prepare Data: Number of MAGs at Family Level ---
MAG_abd_no <- Abs_metadata_MG %>%
  # Keep only MAGs with non-zero abundance
  subset(Relative_abundance > 0) %>%
  
  # Factorize columns for ordering in plots
  mutate(
    `Water type` = factor(`Water type`, levels = c("OSPW", "ROW")),
    `Plant species` = factor(`Plant species`, levels = c("Typha", "Juncus")),
    Time = factor(Time, levels = c("D-0", "D6", "D19", "D40", "D60", "D89"))
  ) %>%
  
  # --- Average Across Replicates ---
  group_by(MAGs, SampleID, `Plant species`, `Water type`, Time, Replicates) %>%
  mutate(DE = mean(Relative_abundance), .groups = "drop") %>%
  
  # --- Count MAGs per Family, Plant, Water Type, and Time ---
  group_by(`Plant species`, `Water type`, Time, Family) %>%
  summarise(count = n_distinct(MAGs), .groups = "drop") 

# --- Plot Number of MAGs at Family Level (Filter count > 5) ---
MAG_abd_no %>%
  filter(count > 5) %>%
  ggplot(aes(x = Family, y = count, fill = `Plant species`)) +
  
  # Bar plot
  geom_bar(stat = "identity", position = position_dodge()) +
  
  # Add counts as text labels
  geom_text(aes(label = count), position = position_dodge(width = 0.9),
            angle = 90, hjust = 0.05, size = 5) +
  
  # Axis labels
  xlab("Plant species") +
  ylab("Number of MAGs") +
  
  # Theme and color settings
  theme_bw() +
  scale_fill_manual(values = plant) +
  
  # Facet by Water Type and Time
  facet_wrap(`Water type` ~ Time, ncol = 3) +
  
  # Customize text and legend
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    strip.text.x = element_text(size = 26),
    strip.text.y = element_text(size = 26),
    legend.text = element_text(colour = "black", size = 24),
    axis.text.x = element_text(
      size = 24, hjust = 1, vjust = 0.5,
      angle = 90, colour = "black"
    ),
    legend.title = element_text(size = 24)
  )

# --- Save Plot ---
ggsave("Figures/Number_of_MAGs_Family_Typha_OSPW_n>5.svg", width = 65, height = 65, units = "cm")

```


**Relative abundance of MAGs at phylum level**
```{r load RT MAGs at phylum, cache = T}
  
# --- Define Colors for Each Phylum ---
phylum_colors <- c(
  "Acidobacteriota" = "#1f77b4", "Actinomycetota" = "#ff7f0e",
  "Armatimonadota" = "#2ca02c", "Ascomycota" = "#ffb6c1",
  "Bacteroidota" = "#9467bd", "Bdellovibrionota" = "#EEC900",
  "Cyanobacteriota" = "#8c564b", "Deinococcota" = "#b22222",
  "Fibrobacterota" = "#ee82ee", "Gemmatimonadota" = "#17becf",
  "Myxococcota" = "#228b22", "Oomycota" = "#a0522d",
  "Planctomycetota" = "#c49c94", "Pseudomonadota" = "#8B0A50",
  "Spirochaetota" = "#98df8a", "Verrucomicrobiota" = "#dbdb8d",
  "Arthropoda" = "#ff4500", "Deferribacterota" = "#ff69b4",
  "Desulfobacterota" = "#ffd700", "Bacillota" = "#d62728"
)

# --- Prepare Data: Relative Abundance at Phylum Level ---
MAG_abd <- Abs_metadata_MG %>%
  # Keep only MAGs with non-zero abundance
  subset(Relative_abundance > 0) %>%
  
  # Factorize for plotting order
  mutate(
    `Water type` = factor(`Water type`, levels = c("OSPW", "ROW")),
    `Plant species` = factor(`Plant species`, levels = c("Typha", "Juncus")),
    Time = factor(Time, levels = c("D-0", "D6", "D19", "D40", "D60", "D89"))
  ) %>%
  
  # --- Average Across Replicates ---
  group_by(MAGs, SampleID, `Plant species`, `Water type`, Time, Replicates) %>%
  mutate(DE = mean(Relative_abundance), .groups = "drop") %>%
  
  # --- Sum Abundance Within Each Phylum ---
  group_by(`Plant species`, `Water type`, Time, Phylum) %>%
  summarise(DE_sum = sum(DE), .groups = "drop") %>%
  
  # --- Calculate Relative Abundance (%) ---
  group_by(`Plant species`, `Water type`, Time) %>%
  mutate(Percent = DE_sum / sum(DE_sum) * 100)

# --- Plot: Relative Abundance of MAGs at Phylum Level ---
MAG_abd %>%
  ggplot(aes(x = `Plant species`, y = Percent, fill = Phylum)) +
  geom_bar(stat = "identity", position = "fill") +
  xlab("Plant species") +
  ylab("Relative abundance (%)") +
  theme_bw() +
  scale_fill_manual(values = phylum_colors) +
  facet_grid(`Water type` ~ Time) +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    strip.text.x = element_text(size = 26),
    strip.text.y = element_text(size = 26),
    legend.text = element_text(colour = "black", size = 24),
    axis.text.x = element_text(
      size = 24, hjust = 1, vjust = 0.5,
      angle = 90, colour = "black"
    ),
    legend.title = element_text(size = 24)
  )

# --- Save Plot ---
ggsave("Figures/Relative_abundance_of_MAGs_Phylum.svg", width = 65, height = 65, units = "cm")

```


**Relative abundance of MAGs at family level**
```{r load RT MAGs at family, cache = T}

# --- Define Colors for Each Family ---
family_colors <- c(
  "Burkholderiaceae" = "#8B2323", "Caulobacteraceae" = "#7FFFD4",
  "Gallionellaceae" = "#E0EEEE", "Chitinophagaceae" = "#00008B",
  "Aestuariivirgaceae" = "#FF4040", "Xanthobacteraceae" = "#8B8378",
  "Methylophilaceae" = "#53868B", "Pseudomonadaceae" = "#7FFF00",
  "Sphingomonadaceae" = "#458B00", "Cellvibrionaceae" = "#8B4513",
  "Opitutaceae" = "#008B8B", "Flavobacteriaceae" = "#FFB90F",
  "Rhodocyclaceae" = "#556B2F", "Cellulomonadaceae" = "#8B4500",
  "Thiobacillaceae" = "#BF3EFF", "Polyangiaceae" = "#68228B",
  "Enterobacteriaceae" = "#B4EEB4", "Denitrovibrionaceae" = "#8B0A50",
  "Acutalibacteraceae" = "#FF1493", "Micropepsaceae" = "#00688B",
  "Lachnospiraceae" = "#8B7500", "Pirellulaceae" = "#FFD700",
  "JABEVD01" = "#008B00", "Acetobacteraceae" = "#00FF00"
)

# --- Prepare Data: Relative Abundance at Family Level ---
MAG_abd <- Abs_metadata_MG %>%
  # Keep only MAGs with non-zero abundance
  subset(Relative_abundance > 0) %>%
  
  # Factorize for plotting order
  mutate(
    `Water type` = factor(`Water type`, levels = c("OSPW", "ROW")),
    `Plant species` = factor(`Plant species`, levels = c("Typha", "Juncus")),
    Time = factor(Time, levels = c("D-0", "D6", "D19", "D40", "D60", "D89"))
  ) %>%
  
  # --- Average Across Replicates ---
  group_by(MAGs, SampleID, `Plant species`, `Water type`, Time, Replicates) %>%
  mutate(DE = mean(Relative_abundance), .groups = "drop") %>%
  
  # --- Sum Abundance Within Each Family ---
  group_by(`Plant species`, `Water type`, Time, Family) %>%
  summarise(DE_sum = sum(DE), .groups = "drop") %>%
  
  # --- Calculate Relative Abundance (%) ---
  group_by(`Plant species`, `Water type`, Time) %>%
  mutate(Percent = DE_sum / sum(DE_sum) * 100) %>%
  
  # --- Keep Top 4 Families per Sample ---
  arrange(`Plant species`, `Water type`, Time, desc(Percent)) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 4) %>%
  ungroup()

# --- Plot: Relative Abundance of Top 4 MAG Families ---
MAG_abd %>%
  ggplot(aes(x = `Plant species`, y = Percent, fill = Family)) +
  geom_bar(stat = "identity", position = "fill") +
  xlab("Plant species") +
  ylab("Relative abundance (%)") +
  theme_bw() +
  scale_fill_manual(values = family_colors) +
  facet_grid(`Water type` ~ Time) +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    strip.text.x = element_text(size = 26),
    strip.text.y = element_text(size = 26),
    legend.text = element_text(colour = "black", size = 24),
    axis.text.x = element_text(
      size = 24, hjust = 1, vjust = 0.5,
      angle = 90, colour = "black"
    ),
    legend.title = element_text(size = 24)
  ) +
  guides(fill = guide_legend(ncol = 1)) 

# --- Save Plot ---
ggsave("Figures/Relative_abundance_of_MAGs_Family_n=4.svg", width = 65, height = 65, units = "cm")

```



**Load MAGs novelty data**
```{r load novelty data, cache = T}
# --- Load Novelty Data from ODS File ---
# This reads the "Novelty" sheet from "bins_summary.ods" and keeps the original column names
read_ods("bins_summary.ods", sheet = "Novelty", col_names = TRUE) ->
  Novelty

```

**Number of novel MAGs**
```{r load n novel MAGs, cache = T}
########### Number of Novel MAGs (Genus Level) ###########

# Filter out eukaryotic MAGs and count MAGs per genus novelty
Novelty %>%
  filter(Domain != "Eukaryota") %>%                     # Exclude eukaryotic MAGs
  group_by(Genus_novelty) %>%                           # Group by genus novelty status
  summarise(count = n(), .groups = "drop") %>%          # Count number of MAGs per group
  mutate(Genus_novelty = factor(Genus_novelty,          # Ensure ordering of categories
                                levels = c("Known", "Novel"))) %>%
  
# Plot the number of MAGs per genus novelty
ggplot(aes(x = Genus_novelty, y = count)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Bar plot
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            hjust = 0.05, vjust = 0.05, size = 8) +          # Add counts on bars
  ylab("Number of MAGs") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "none"                                # Hide legend
  )

# Save plot
ggsave("Figures/Number_of_MAGs_Genus_novelty.svg", 
       width = 45, height = 45, units = "cm")

```

**Number of species-level novelty**
```{r load species-level novelty, cache = T}
########### Number of Novel MAGs (Species Level) ###########

# Filter out eukaryotic MAGs and count MAGs per species novelty
Novelty %>%
  filter(Domain != "Eukaryota") %>%                     # Exclude eukaryotic MAGs
  group_by(Species_novelty) %>%                         # Group by species novelty status
  summarise(count = n(), .groups = "drop") %>%          # Count number of MAGs per group
  mutate(Species_novelty = factor(Species_novelty,      # Ensure ordering of categories
                                  levels = c("Known", "Novel"))) %>%
  
# Plot the number of MAGs per species novelty
ggplot(aes(x = Species_novelty, y = count)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Bar plot
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            hjust = 0.05, vjust = 0.05, size = 8) +          # Add counts on bars
  ylab("Number of MAGs") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "none"                                # Hide legend
  )

# Save plot
ggsave("Figures/Number_of_MAGs_Species_novelty.svg", 
       width = 45, height = 45, units = "cm")

```


**The phylum of species-level of novel MAGs**
```{r load phylum novel MAGs, cache = T}

########### Phylum distribution of species-level novel MAGs ###########

# Define phylum colors
phylum_colors <- c(
  "Acidobacteriota" = "#1f77b4", "Actinomycetota" = "#ff7f0e",
  "Armatimonadota" = "#2ca02c", "Ascomycota" = "#ffb6c1",
  "Bacteroidota" = "#9467bd", "Bdellovibrionota" = "#EEC900",
  "Cyanobacteriota"  = "#8c564b", "Deinococcota" = "#b22222",
  "Fibrobacterota" = "#ee82ee", "Gemmatimonadota" = "#17becf",
  "Myxococcota"  = "#228b22", "Oomycota" = "#a0522d",
  "Planctomycetota" = "#c49c94", "Pseudomonadota" = "#8B0A50",
  "Spirochaetota" = "#98df8a", "Verrucomicrobiota" = "#dbdb8d",
  "Arthropoda" = "#ff4500", "Deferribacterota" = "#ff69b4",
  "Desulfobacterota" = "#ffd700", "Bacillota" = "#d62728"
)

# Prepare data and plot
Novelty %>%
  filter(Domain != "Eukaryota") %>%                       # Exclude eukaryotic MAGs
  group_by(Species_novelty, Phylum) %>%                   # Group by novelty status and phylum
  summarise(count = n(), .groups = "drop") %>%            # Count MAGs in each group
  mutate(Species_novelty = factor(Species_novelty,        # Order novelty categories
                                  levels = c("Known", "Novel"))) %>%
  
  # Plot
  ggplot(aes(x = Species_novelty, y = count, fill = Phylum)) +
  geom_bar(stat = "identity", position = "dodge") +       # Side-by-side bars
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            hjust = 0.5, vjust = 0.05, size = 8) +       # Counts on top of bars
  ylab("Number of MAGs") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    legend.text = element_text(colour = "black", size = 24),
    axis.text.x = element_text(size = 24, hjust = 0.5, vjust = 0.5, colour = "black"),
    legend.title = element_text(size = 24)
  ) +
  guides(fill = guide_legend(ncol = 1)) +                # Single-column legend
  scale_fill_manual(values = phylum_colors)              # Apply phylum colors

# Save figure
ggsave("Figures/Number_of_MAGs_Species_novelty_Phylum.svg", width = 45, height = 45, units = "cm")

```

**The family of species-level of novel MAGs**
```{r load family novel_MAGs, cache = T}
########### Family distribution of species-level novel MAGs ###########

# Define family colors
family_colors <- c(
  "Burkholderiaceae" = "#8B2323", "Caulobacteraceae" = "#7FFFD4",
  "Gallionellaceae" = "#E0EEEE", "Chitinophagaceae" = "#00008B",
  "Aestuariivirgaceae" = "#FF4040", "Xanthobacteraceae" = "#8B8378",
  "Methylophilaceae" = "#53868B", "Pseudomonadaceae" = "#7FFF00",
  "Sphingomonadaceae" = "#458B00", "Cellvibrionaceae" = "#8B4513",
  "Opitutaceae" = "#008B8B", "Flavobacteriaceae" = "#FFB90F",
  "Rhodocyclaceae" = "#556B2F", "Cellulomonadaceae" = "#8B4500",
  "Thiobacillaceae" = "#BF3EFF", "Polyangiaceae" = "#68228B",
  "Enterobacteriaceae" = "#B4EEB4", "Denitrovibrionaceae" = "#8B0A50",
  "Acutalibacteraceae" = "#FF1493", "Micropepsaceae" = "#00688B",
  "Lachnospiraceae" = "#8B7500", "Pirellulaceae" = "#FFD700",
  "JABEVD01" = "#008B00", "Acetobacteraceae" = "#00FF00",
  "B-17BO" = "#FAEBD7", "Beijerinckiaceae" = "#76EEC6",
  "Bryobacteraceae" = "#5F9EA0", "Gemmataceae" = "#D2691E",
  "Planctomycetaceae" = "#A2CD5A", "Rhizobiaceae" = "#8DEEEE",
  "Steroidobacteraceae" = "#EEEE00", "Tepidisphaeraceae" = "#FFE1FF"
)

# Prepare data: exclude eukaryotes, group by novelty and family, filter rare families
Novelty %>%
  filter(Domain != "Eukaryota") %>%                        
  group_by(Species_novelty, Family) %>%                     
  summarise(count = n(), .groups = "drop") %>%             
  mutate(Species_novelty = factor(Species_novelty, levels = c("Known", "Novel"))) %>% 
  filter(count > 3) -> nol_3                                  # Keep families with >3 MAGs

# Plot
nol_3 %>%
  ggplot(aes(x = Species_novelty, y = count, fill = Family)) +
  geom_bar(stat = "identity", position = "dodge") +         # Side-by-side bars
  geom_text(aes(label = count), 
            position = position_dodge(width = 0.9), 
            hjust = 0.5, vjust = 0.05, size = 8) +         # Count labels
  ylab("Number of MAGs") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 24, colour = "black"),
    axis.ticks = element_line(size = 0.8),
    axis.title = element_text(size = 24),
    legend.position = "right",
    legend.text = element_text(colour = "black", size = 24),
    axis.text.x = element_text(size = 24, hjust = 0.5, vjust = 0.5, colour = "black"),
    legend.title = element_text(size = 24)
  ) +
  guides(fill = guide_legend(ncol = 1)) +                   # Single-column legend
  scale_fill_manual(values = family_colors)                 # Apply custom colors

# Save figure
ggsave("Figures/Number_of_MAGs_Species_novelty_Family.svg", width = 45, height = 45, units = "cm")

```
